---
title: "DAssemble: Ensemble Differential Analysis"
author:
  - name: Himel Mallick
  - name: Ziyu Liu
  - name: Nalin Arora
package: DAssemble
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{DAssemble: Ensemble Differential Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", warning = FALSE
)
```

# Introduction
DAssemble is an R package for ensemble-based differential association analysis 
of high-throughput omics data. It provides a flexible framework for combining 
a user-specified **core method** with one or more complementary **enhancer** methods, 
allowing users to integrate statistical evidence across different modeling assumptions 
rather than relying on a single differential analysis approach. DAssemble aggregates 
model-specific p-values using the **Cauchy Combination Tests (CCT)** and reports both 
ensemble-level and method-specific results.

The package was developed to support robust and interpretable differential analysis 
across multiple omics domains, including bulk RNA-seq, single-cell RNA-seq, and 
microbiome data. We submitted DAssemble to Bioconductor to make the method easier 
to install, document, test, and integrate with established Bioconductor workflows 
and data structures, thereby supporting reproducible and extensible analysis of 
high-throughput biological data.

# Overview

`DAssemble` implements an ensemble framework for differential-abundance and
differential-expression analysis. Users can choose a single core model and
optionally combine it with enhancer tests using the Cauchy Combination Test.

The package accepts a `MultiAssayExperiment` object directly. When multiple
experiments are present, use `assay_name` to select the experiment to analyze.
For lower-level workflows, `features` can also be a data frame with samples in
rows and features in columns, paired with a `metadata` data frame with matching
sample row names. The exposure variable supplied through `expVar` must be
binary.

The main `DAssemble()` function also supports:

- multiple adjustment covariates through `coVars`
- repeated-measures / longitudinal designs through `random_effects`
- open-ended method-specific controls through `method_args`

These arguments are available only for methods whose wrappers can express the
corresponding model directly. For unsupported methods, DAssemble stops with a
clear error rather than fitting a mis-specified model.

# Main API

The primary entry point is:

```{r main-signature, eval=FALSE}
DAssemble(
  features,
  metadata = NULL,
  core_method = NULL,
  enhancers = NULL,
  expVar = "group",
  coVars = NULL,
  random_effects = NULL,
  method_args = NULL,
  assay_name = NULL,
  p_adj = "BY",
  enhancer_norm = "TSS",
  return_components = TRUE,
  return_subensembles = FALSE
)
```

`method_args` is a named control list used to forward additional arguments to
underlying methods. The usual pattern is:

- `method_args$core` for arguments shared by the selected core method
- `method_args$enhancer` for arguments shared by the selected enhancers
- `method_args$<MethodName>` for method-specific overrides

For the `Prevalence` core, the wrapper additionally recognizes
`method_args$Prevalence$separation_method`. The default is `"augment"`,
which follows the MaAsLin3-style augmented logistic fit. For
cross-sectional prevalence analyses, users can instead request
`"firth"` to use bias-reduced logistic regression through `brglm2`.

Wrappers with multi-stage internals can also accept subcall-specific entries.
For example, `method_args$DESeq2` may contain items such as `DESeq`,
`results`, or `estimateSizeFactors`, while `method_args$edgeR` may contain
items such as `glmFit` or `glmLRT`.

```{r method-args-example, eval=FALSE}
res <- DAssemble(
  features = features,
  metadata = metadata,
  core_method = "Prevalence",
  expVar = "group",
  method_args = list(
    Prevalence = list(separation_method = "firth")
  )
)
```

# Supported methods

Core methods currently available in `DAssemble()` include:

- `Prevalence`
- `DESeq2`
- `edgeR`
- `limmaVOOM`
- `metagenomeSeq`
- `MAST`
- `dearseq`
- `ALDEx2`
- `LinDA`
- `LOCOM`
- `Maaslin2`
- `Maaslin3`
- `Tweedieverse`
- `Robseq`
- `ANCOMBC2`

Enhancers currently available are:

- `WLX`
- `LR`
- `KS`

# Covariate and longitudinal support

The current implementation supports multiple covariates (`coVars`) for the
following methods:

| Method | Multiple covariates |
|:--|:--|
| Prevalence | Yes |
| DESeq2 | Yes |
| edgeR | Yes |
| limmaVOOM | Yes |
| metagenomeSeq | Yes |
| MAST | Yes |
| dearseq | Yes |
| ALDEx2 | Yes |
| LinDA | Yes |
| Maaslin2 | Yes |
| Maaslin3 | Yes |
| Tweedieverse | Yes |
| Robseq | Yes |
| ANCOMBC2 | Yes |
| LR | Yes |
| WLX | No |
| KS | No |
| LOCOM | No |

When `random_effects` is supplied, the following methods are currently treated
as longitudinal-compatible:

| Method | Longitudinal support |
|:--|:--|
| Prevalence | Yes |
| Maaslin2 | Yes |
| Maaslin3 | Yes |
| Tweedieverse | Yes |
| LR | Yes |

All other methods should be treated as cross-sectional only in the current
package implementation.

For the `Prevalence` core, the default fitting path uses MaAsLin3-style
augmentation in both cross-sectional and longitudinal settings. The
alternative `method_args$Prevalence$separation_method = "firth"` is
available only for the non-longitudinal prevalence model.

`Maaslin3` is used through its standard package pathway in `DAssemble()` and
is not forced into a prevalence-only mode.

# Installation

Install `DAssemble` from Bioconductor with `BiocManager`:

```{r installation, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install("DAssemble")
```

# Bulk RNA-seq Example

The `airway` dataset from Bioconductor contains RNA-seq counts from airway
smooth muscle cells treated with dexamethasone. The example below runs
`DAssemble` with `DESeq2` as the core method and two enhancers.

```{r airway-example}
data("airway", package = "airway")
counts <- SummarizedExperiment::assay(airway, "counts")
keep_genes <- order(rowSums(counts), decreasing = TRUE)[seq_len(500L)]
counts <- counts[keep_genes, , drop = FALSE]
metadata <- as.data.frame(SummarizedExperiment::colData(airway))
metadata <- metadata[colnames(counts), , drop = FALSE]

se <- SummarizedExperiment::SummarizedExperiment(
  assays = list(counts = counts)
)

mae <- MultiAssayExperiment::MultiAssayExperiment(
  experiments = list(rnaseq = se),
  colData = S4Vectors::DataFrame(metadata)
)

res <- DAssemble::DAssemble(
  features = mae,
  assay_name = "rnaseq",
  core_method = "DESeq2",
  enhancers = c("WLX", "LR"),
  expVar = "dex",
  coVars = c("cell", "albut"),
  p_adj = "BH",
  enhancer_norm = "tmm",
  method_args = list(
    DESeq2 = list(
      DESeq = list(fitType = "local")
    ),
    LR = list(
      control = glm.control(maxit = 100)
    )
  ),
  return_components = TRUE,
  return_subensembles = TRUE
)

head(res$res)
names(res$components)
```

This analysis adjusts for `cell` and `albut`, combines DESeq2 p-values with
Wilcoxon and logistic-regression enhancer p-values, and forwards example
control arguments through `method_args`. The combined results are returned in
`res$res`, and the per-method outputs are available in `res$components`.

# Longitudinal example

The example below illustrates a longitudinal-style call using a
longitudinal-compatible core method and the `LR` enhancer.

```{r longitudinal-example, eval=FALSE}
res <- DAssemble::DAssemble(
  features = features,
  metadata = metadata,
  core_method = "Maaslin3",
  enhancers = "LR",
  expVar = "group",
  coVars = c("age", "sex", "batch"),
  random_effects = "subject",
  method_args = list(
    Maaslin3 = list(max_significance = 0.25),
    LR = list(control = glm.control(maxit = 100))
  ),
  p_adj = "BH"
)
```

# Microbiome Example

The `GlobalPatterns` dataset from `phyloseq` contains 16S rRNA profiles from
environmental and host-associated microbiome samples. The example below compares
fecal and soil samples using enhancer-only DAssemble with CLR normalization.

```{r microbiome-example}
data("GlobalPatterns", package = "phyloseq")
gp <- GlobalPatterns

otu <- as(phyloseq::otu_table(gp), "matrix")
metadata <- as.data.frame(phyloseq::sample_data(gp))

keep <- metadata$SampleType %in% c("Feces", "Soil")
otu <- otu[, keep, drop = FALSE]
keep_taxa <- order(rowSums(otu), decreasing = TRUE)[seq_len(250L)]
features <- as.data.frame(t(otu[keep_taxa, , drop = FALSE]))
metadata <- metadata[keep, , drop = FALSE]
metadata$group <- droplevels(factor(metadata$SampleType))
stopifnot(nlevels(metadata$group) == 2L)

se <- SummarizedExperiment::SummarizedExperiment(
  assays = list(counts = t(as.matrix(features)))
)

mae <- MultiAssayExperiment::MultiAssayExperiment(
  experiments = list(microbiome = se),
  colData = S4Vectors::DataFrame(metadata)
)

res <- DAssemble::DAssemble(
  features = mae,
  assay_name = "microbiome",
  core_method = NULL,
  enhancers = c("WLX", "LR", "KS"),
  expVar = "group",
  enhancer_norm = "clr",
  return_components = TRUE,
  return_subensembles = TRUE
)

head(res$res)
names(res$components)
head(res$ensembles)
```

Because the core method is `NULL`, this example combines only enhancer tests.
The `$ensembles` element reports the CCT results for the enhancer combinations
when `return_subensembles = TRUE`.

# Single-Cell Example

The single-cell example uses a small `SingleCellExperiment` object with two cell
groups. The object is wrapped in a `MultiAssayExperiment` before being passed to
`DAssemble`, matching the same input architecture used above.

```{r single-cell-example}
set.seed(1)
n_genes <- 120L
n_cells <- 20L

cell_type <- factor(rep(c("control", "treated"), each = n_cells / 2L))
counts <- matrix(
  stats::rnbinom(n_genes * n_cells, mu = 20, size = 5),
  nrow = n_genes,
  dimnames = list(
    paste0("gene", seq_len(n_genes)),
    paste0("cell", seq_len(n_cells))
  )
)

counts[seq_len(12L), cell_type == "treated"] <-
  counts[seq_len(12L), cell_type == "treated"] + 15L

metadata <- data.frame(
  CellType = cell_type,
  row.names = colnames(counts)
)

sce <- SingleCellExperiment::SingleCellExperiment(
  assays = list(counts = counts),
  colData = S4Vectors::DataFrame(metadata)
)

mae <- MultiAssayExperiment::MultiAssayExperiment(
  experiments = list(single_cell = sce),
  colData = S4Vectors::DataFrame(metadata)
)

res <- DAssemble::DAssemble(
  features = mae,
  assay_name = "single_cell",
  core_method = "edgeR",
  enhancers = c("WLX", "LR", "KS"),
  enhancer_norm = "TSS",
  expVar = "CellType",
  p_adj = "BH",
  return_components = TRUE,
  return_subensembles = TRUE
)

head(res$res)
names(res$components)
```
# Session information

```{r session-info}
sessionInfo()
```
