---
title: "DAssemble: Ensemble Differential Analysis"
author:
  - name: Himel Mallick
  - name: Ziyu Liu
  - name: Nalin Arora
package: DAssemble
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{DAssemble: Ensemble Differential Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", warning = F
)
```

# Overview

`DAssemble` implements an ensemble framework for differential-abundance and
differential-expression analysis. Users can choose a single core model and
optionally combine it with enhancer tests using the Cauchy Combination Test.

The package accepts a `MultiAssayExperiment` object directly. When multiple
experiments are present, use `assay_name` to select the experiment to analyze.
For lower-level workflows, `features` can also be a data frame with samples in
rows and features in columns, paired with a `metadata` data frame with matching
sample row names. The exposure variable supplied through `expVar` must be
binary.

# Installation

Install `DAssemble` from Bioconductor with `BiocManager`:

```{r installation, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install("DAssemble")
```

# Bulk RNA-seq Example

The `airway` dataset from Bioconductor contains RNA-seq counts from airway
smooth muscle cells treated with dexamethasone. The example below runs
`DAssemble` with `DESeq2` as the core method and two enhancers.

```{r airway-example}
example_pkgs <- c("airway", "DESeq2", "MultiAssayExperiment", "S4Vectors")
if (all(vapply(example_pkgs, requireNamespace, logical(1), quietly = TRUE))) {
  data("airway", package = "airway")
  counts <- SummarizedExperiment::assay(airway, "counts")
  metadata <- as.data.frame(SummarizedExperiment::colData(airway))
  metadata <- metadata[colnames(counts), , drop = FALSE]

  mae <- MultiAssayExperiment::MultiAssayExperiment(
    experiments = list(rnaseq = SummarizedExperiment::SummarizedExperiment(
      assays = list(counts = counts)
    )),
    colData = S4Vectors::DataFrame(metadata)
  )

  res <- DAssemble::DAssemble(
    features = mae,
    assay_name = "rnaseq",
    core_method = "DESeq2",
    enhancers = c("WLX", "LR"),
    expVar = "dex",
    p_adj = "BH",
    enhancer_norm = "tmm",
    return_components = TRUE,
    return_subensembles = TRUE
  )

  head(res$res)
  names(res$components)
}
```

This analysis combines DESeq2 p-values with Wilcoxon and logistic-regression
enhancer p-values. The combined results are returned in `res$res`, and the
per-method outputs are available in `res$components`.

# Microbiome Example

The `GlobalPatterns` dataset from `phyloseq` contains 16S rRNA profiles from
environmental and host-associated microbiome samples. The example below compares
fecal and soil samples using enhancer-only DAssemble with CLR normalization.

```{r microbiome-example}
example_pkgs <- c("phyloseq", "MultiAssayExperiment", "S4Vectors")
if (all(vapply(example_pkgs, requireNamespace, logical(1), quietly = TRUE))) {
  data("GlobalPatterns", package = "phyloseq")
  gp <- GlobalPatterns

  otu <- as(phyloseq::otu_table(gp), "matrix")
  metadata <- as.data.frame(phyloseq::sample_data(gp))

  keep <- metadata$SampleType %in% c("Feces", "Soil")
  features <- as.data.frame(t(otu[, keep]))
  metadata <- metadata[keep, , drop = FALSE]
  metadata$group <- droplevels(factor(metadata$SampleType))
  stopifnot(nlevels(metadata$group) == 2L)

  mae <- MultiAssayExperiment::MultiAssayExperiment(
    experiments = list(microbiome = SummarizedExperiment::SummarizedExperiment(
      assays = list(counts = t(as.matrix(features)))
    )),
    colData = S4Vectors::DataFrame(metadata)
  )

  res <- DAssemble::DAssemble(
    features = mae,
    assay_name = "microbiome",
    core_method = NULL,
    enhancers = c("WLX", "LR", "KS"),
    expVar = "group",
    enhancer_norm = "clr",
    return_components = TRUE,
    return_subensembles = TRUE
  )

  head(res$res)
  names(res$components)
  head(res$ensembles)
}
```

Because the core method is `NULL`, this example combines only enhancer tests.
The `$ensembles` element reports the CCT results for the enhancer combinations
when `return_subensembles = TRUE`.

# Session information

```{r session-info}
sessionInfo()
```
