---
title: "Case study 3: Interpretable ensemble models for differential expression analysis of Microbiome data"
author: "Jialin Gao"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tweedieverse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette highlights the example workflow for performing differential expression using interpretable ensemble models with a microbiome dataset.

## Load Dataset
To install Github package, users need to create their own GitHub Personal Access Token (PAT). You can follow this [`guide`](https://docs.github.com/en/authentication/keeping-your-account-and-data-secure/managing-your-personal-access-tokens) if you have trouble creating one.
```{r, warning=FALSE, message = FALSE}
# Clear workspace
rm(list = ls())

# Load libraries
library(DAssemble)
library(tidyverse)
library(Maaslin2)
library(data.table)
library(SummarizedExperiment)
library(Biobase)
library(scatterplot3d)
library(withr)
library(gridExtra)
library(LOCOM) # devtools::install_github("yijuanhu/LOCOM")
library(BiocParallel)
library(ggplot2)
library(cowplot)
library(dplyr)
#install.packages("here")
library(here)
# Load Source Files
setwd(here::here())
pkgmaker::source_files('./Codes/Real_Data/helper', '*.R')
```

## Data Loading and Preprocessing
```{r}
file_path <- "~/Desktop/DAssemble/Codes/Dataset/HMP2_features_baseline_counts.txt"
features <- read.table(file_path, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
metadata <- read_tsv('~/Desktop/DAssemble/Codes/Dataset/HMP2_metadata.tsv')


input_features <- as.data.frame(features)
rownames(input_features) <- input_features$ID
input_features <- input_features[-1] # Remove ID column


input_metadata <- as.data.frame(metadata)
rownames(input_metadata) <- metadata$ID

input_metadata <- na.omit(input_metadata)
input_metadata <- input_metadata[rownames(input_metadata) %in% rownames(input_features), ]


# Assign 1 for IBD (both UC and CD) and 0 for nonIBD
input_metadata$diagnosis <- ifelse(input_metadata$diagnosis %in% c("UC", "CD"), 1, 0)
input_metadata$antibiotics <- ifelse(input_metadata$antibiotics == "No", 0, 1)
input_features <- input_features[row.names(input_features) %in% row.names(input_metadata), ]

```

## Output
Here, we fit the two part model using MaAsLin2 and LOCOM( A logistic regression model for testing differential abundance in compositional microbiome data). First, we fit the models separately and then combine the two-part model using the p-value combination method CCT.

### Run MaAsLin2
```{r}
## MaAslin2
fit.Maaslin2 <- function(features, metadata, libSize) {
    #############################
    # Standard MaAslin2 pipeline #
    #############################

    # Ensure features and metadata are data frames
    features <- as.data.frame(features)
    metadata <- as.data.frame(metadata)

    # Run MaAslin2
    fit_data <- Maaslin2(input_data = features,
                         min_abundance = - Inf,
                         input_metadata = metadata,
                         output = "Maaslin2_output",
                         normalization = "TSS",
                         transform = "LOG",
                         fixed_effects = c("diagnosis","age","antibiotics"))

    res_df <- fit_data$results
    # Extract coefficients and p-values
    coef <- res_df$coef
    pval_CR <- res_df$pval
    paras <- data.frame(coef_CR = coef, pval_CR = pval_CR)
    paras$feature <- res_df$feature


    #################
    # Return output #
    #################

    paras <- dplyr::select(paras, c('feature'), everything())
    paras$metadata <- res_df$name
    rownames(paras) <- NULL
    return(paras)
}

libSize <- rowSums(input_features)

paras_CR <- fit.Maaslin2(input_features, input_metadata, libSize)
# Filter results to only include entries related to 'diagnosis'
paras_CR_diagnosis <- paras_CR[paras_CR$metadata == "diagnosis", ]
print(paras_CR_diagnosis)
```

### Run LOCOM
```{r}
fit.LOCOM <- function(features, metadata, libSize, ID){
  
  #############
  # Run LOCOM #
  #############
  otu.table <- as.matrix(input_features)
  Y <- as.numeric(input_metadata[, c("diagnosis")])
  C <- data.frame(age = as.numeric(input_metadata[, "age"]), 
                      antibiotics = as.numeric(input_metadata[, "antibiotics"]))
  
  filter.threshold <- 0
  res <- locom(otu.table = otu.table, Y = Y, C = C, fdr.nominal = 0.1, seed = 1, n.perm.max = 10000,  filter.thresh = filter.threshold)
  
  paras <- data.frame(pval_LO = res$p.otu[1,], coef_LO = res$effect.size[1,])

  paras$feature <- rownames(paras)
  paras$metadata <- "diagnosis"
  
  #################
  # Return output #
  #################
  
  paras <- paras[order(paras$pval_LO, decreasing = FALSE),]
  paras <- dplyr::select(paras, c('feature'), everything())
  
  rownames(paras) <- NULL
  return(paras) 
}

paras_LO <- fit.LOCOM(input_features, input_metadata, libSize, ID)
paras_LO

```

### Combine results
```{r}
paras_CR <- paras_CR_diagnosis %>%
  dplyr::mutate(ID = feature) %>%  
  dplyr::rename(pvalue = pval_CR) 

paras_LO = paras_LO %>%
  dplyr::mutate(ID = feature) %>%  
  dplyr::rename(pvalue = pval_LO)

# Combine the results
dflist = list(paras_LO, paras_CR)
```

After running MaAsLin2 and LOCOM, we use Vanilla, Cauchy Combination Test (CCT) and stouffer to combine p-values and calculate q-values based on combined p-values. We can use the DAssemble() from library(DAssemble).

#### Vanilla
```{r}
# P-value: Vanilla
parasVanilla = DAssemble(dflist, combine.method = "minP", correction = "BH")
parasVanilla[1:10, ]
```
#### CCT
```{r, warning=FALSE}
# P-value: CCT
parasCCT = DAssemble(dflist, combine.method = "CCT", correction = "BH")
parasCCT[1:10,]
```
#### stouffer
```{r, warning=FALSE}
# P-value: stouffer
parasStouffer = DAssemble(dflist, combine.method = "stouffer", correction = "BH")
parasStouffer[1:10,]
```


## Results Comparison 

### LOCOM vs MaAsLin2(CCT)
```{r, fig.width=7, fig.height=5.5, warning=FALSE}
# Rank features based on q-value and select top 10 features
top_10 <- parasCCT %>%
  arrange(qval_BH) %>%
  slice_head(n = 10) %>%
  mutate(
    feature = coalesce(feature.x, feature.y), 
    pval_CR_log10 = -log10(pvalue2),          # From MaAsLin2 results
    pval_LO_log10 = -log10(pvalue1)           # From LOCOM results
  )

pval_range <- range(c(top_10$pval_CR_log10, abs(top_10$pval_LO_log10)))
breaks_pretty_pval <- pretty(c(-max(pval_range), max(pval_range)), n = 5)

pyramid_plot_pval <- ggplot(top_10) +
  geom_col(aes(x = feature, y = pval_CR_log10, fill = "MaAsLin2"), position = "identity") +
  geom_col(aes(x = feature, y = -pval_LO_log10, fill = "LOCOM"), position = "identity") +
  labs(y = "-log10(p-value)", x = "") +  
  coord_flip() +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.y = element_text(size = 10, face = "italic"),
    axis.text.x = element_text(size = 10)
  ) +
  scale_fill_manual(values = c("MaAsLin2" = "darkblue", "LOCOM" = "orange")) +
  scale_y_continuous(breaks = breaks_pretty_pval, labels = abs(breaks_pretty_pval), limits = range(breaks_pretty_pval))

coef_range <- range(c(top_10$coef_CR, abs(top_10$coef_LO)))
breaks_pretty_coef <- pretty(c(-max(coef_range), max(coef_range)), n = 5)

pyramid_plot_coef <- ggplot(top_10) +
  geom_col(aes(x = feature, y = abs(coef_CR), fill = "MaAsLin2"), position = "identity") +
  geom_col(aes(x = feature, y = -abs(coef_LO), fill = "LOCOM"), position = "identity") +
  geom_text(aes(x = feature, y = abs(coef_CR), label = ifelse(coef_CR > 0, "+", "-"), hjust = -0.3), size = 4) +
geom_text(aes(x = feature, y = -abs(coef_LO), label = ifelse(coef_LO > 0, "+", "-"), hjust = 1.3), size = 4) +
  labs(y = "Coefficient", fill = "") +
  coord_flip() +
  theme_minimal() +
  theme(
    legend.position = "right",
    axis.title.y = element_blank(), 
    axis.text.y = element_blank(), 
    axis.ticks.y = element_blank(), 
    plot.title = element_blank()
  ) +
  scale_fill_manual(values = c("MaAsLin2" = "darkblue", "LOCOM" = "orange")) +
  scale_y_continuous(breaks = breaks_pretty_coef, labels = abs(breaks_pretty_coef), limits = range(breaks_pretty_coef))

combined_plot <- plot_grid(
  pyramid_plot_pval,
  pyramid_plot_coef,
  ncol = 2,
  align = "h"
)

final_plot <- plot_grid(
  ggdraw() + draw_label(
    "", 
    x = 0.5, hjust = 0.5, size = 13
  ),
  combined_plot,
  ncol = 1,
  rel_heights = c(0.1, 1)
)

print(final_plot)

```

### LOCOM vs MaAsLin2(Stouffer)
```{r, fig.width=7, fig.height=5.5, warning=FALSE}
top_10 <- parasStouffer %>%
  arrange(qval_BH) %>%
  slice_head(n = 10) %>%
  mutate(
    feature = coalesce(feature.x, feature.y), 
    pval_CR_log10 = -log10(pvalue2),          # From MaAsLin2 results
    pval_LO_log10 = -log10(pvalue1)           # From LOCOM results
  )

pval_range <- range(c(top_10$pval_CR_log10, abs(top_10$pval_LO_log10)))
breaks_pretty_pval <- pretty(c(-max(pval_range), max(pval_range)), n = 5)

pyramid_plot_pval <- ggplot(top_10) +
  geom_col(aes(x = feature, y = pval_CR_log10, fill = "MaAsLin2"), position = "identity") +
  geom_col(aes(x = feature, y = -pval_LO_log10, fill = "LOCOM"), position = "identity") +
  labs(y = "-log10(p-value)", x = "") +  
  coord_flip() +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.y = element_text(size = 10, face = "italic"),
    axis.text.x = element_text(size = 10)
  ) +
  scale_fill_manual(values = c("MaAsLin2" = "darkblue", "LOCOM" = "orange")) +
  scale_y_continuous(breaks = breaks_pretty_pval, labels = abs(breaks_pretty_pval), limits = range(breaks_pretty_pval))

coef_range <- range(c(top_10$coef_CR, abs(top_10$coef_LO)))
breaks_pretty_coef <- pretty(c(-max(coef_range), max(coef_range)), n = 5)

pyramid_plot_coef <- ggplot(top_10) +
  geom_col(aes(x = feature, y = abs(coef_CR), fill = "MaAsLin2"), position = "identity") +
  geom_col(aes(x = feature, y = -abs(coef_LO), fill = "LOCOM"), position = "identity") +
  geom_text(aes(x = feature, y = abs(coef_CR), label = ifelse(coef_CR > 0, "+", "-"), hjust = -0.3), size = 4) +
geom_text(aes(x = feature, y = -abs(coef_LO), label = ifelse(coef_LO > 0, "+", "-"), hjust = 1.3), size = 4) +
  labs(y = "Coefficient", fill = "") +
  coord_flip() +
  theme_minimal() +
  theme(
    legend.position = "right",
    axis.title.y = element_blank(), 
    axis.text.y = element_blank(), 
    axis.ticks.y = element_blank(), 
    plot.title = element_blank()
  ) +
  scale_fill_manual(values = c("MaAsLin2" = "darkblue", "LOCOM" = "orange")) +
  scale_y_continuous(breaks = breaks_pretty_coef, labels = abs(breaks_pretty_coef), limits = range(breaks_pretty_coef))

combined_plot <- plot_grid(
  pyramid_plot_pval,
  pyramid_plot_coef,
  ncol = 2,
  align = "h"
)

final_plot <- plot_grid(
  ggdraw() + draw_label(
    "", 
    x = 0.5, hjust = 0.5, size = 13
  ),
  combined_plot,
  ncol = 1,
  rel_heights = c(0.1, 1)
)

print(final_plot)
```


### LOCOM vs MaAsLin2(Vanilla_Combine)
```{r, fig.width=7, fig.height=5.5, warning=FALSE}
top_10 <- parasVanilla %>%
  arrange(qval_BH) %>%
  slice_head(n = 10) %>%
  mutate(
    feature = coalesce(feature.x, feature.y),  
    pval_CR_log10 = -log10(pvalue2),          # From MaAsLin2 results
    pval_LO_log10 = -log10(pvalue1)           # From LOCOM results
  )

pval_range <- range(c(top_10$pval_CR_log10, abs(top_10$pval_LO_log10)))
breaks_pretty_pval <- pretty(c(-max(pval_range), max(pval_range)), n = 5)

pyramid_plot_pval <- ggplot(top_10) +
  geom_col(aes(x = feature, y = pval_CR_log10, fill = "MaAsLin2"), position = "identity") +
  geom_col(aes(x = feature, y = -pval_LO_log10, fill = "LOCOM"), position = "identity") +
  labs(y = "-log10(p-value)", x = "") +  
  coord_flip() +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.y = element_text(size = 10, face = "italic"),
    axis.text.x = element_text(size = 10)
  ) +
  scale_fill_manual(values = c("MaAsLin2" = "darkblue", "LOCOM" = "orange")) +
  scale_y_continuous(breaks = breaks_pretty_pval, labels = abs(breaks_pretty_pval), limits = range(breaks_pretty_pval))

coef_range <- range(c(top_10$coef_CR, abs(top_10$coef_LO)))
breaks_pretty_coef <- pretty(c(-max(coef_range), max(coef_range)), n = 5)

pyramid_plot_coef <- ggplot(top_10) +
  geom_col(aes(x = feature, y = abs(coef_CR), fill = "MaAsLin2"), position = "identity") +
  geom_col(aes(x = feature, y = -abs(coef_LO), fill = "LOCOM"), position = "identity") +
  geom_text(aes(x = feature, y = abs(coef_CR), label = ifelse(coef_CR > 0, "+", "-"), hjust = -0.3), size = 4) +
geom_text(aes(x = feature, y = -abs(coef_LO), label = ifelse(coef_LO > 0, "+", "-"), hjust = 1.3), size = 4) +
  labs(y = "Coefficient", fill = "") +
  coord_flip() +
  theme_minimal() +
  theme(
    legend.position = "right",
    axis.title.y = element_blank(), 
    axis.text.y = element_blank(), 
    axis.ticks.y = element_blank(), 
    plot.title = element_blank()
  ) +
  scale_fill_manual(values = c("MaAsLin2" = "darkblue", "LOCOM" = "orange")) +
  scale_y_continuous(breaks = breaks_pretty_coef, labels = abs(breaks_pretty_coef), limits = range(breaks_pretty_coef))

combined_plot <- plot_grid(
  pyramid_plot_pval,
  pyramid_plot_coef,
  ncol = 2,
  align = "h"
)

final_plot <- plot_grid(
  ggdraw() + draw_label(
    "", 
    x = 0.5, hjust = 0.5, size = 13
  ),
  combined_plot,
  ncol = 1,
  rel_heights = c(0.1, 1)
)

print(final_plot)
```

## Session information
```{r}
sessionInfo()
```

