---
title: 'Case study 3: Interpretable ensemble models for differential expression analysis
  of Microbiome data'
author: "Jialin Gao"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tweedieverse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette highlights the example workflow for performing differential expression using interpretable ensemble models with a microbiome dataset.

## Install and load packages
```{r, warning=FALSE, message = FALSE}
# Clear workspace
rm(list = ls())
#install.packages(c("lme4", "dplyr"))
library(lme4)
library(dplyr)
library(readr)
library(foreach)
library(doParallel)
library(Maaslin2)
library(ANCOMBC)
library(TreeSummarizedExperiment)
library(S4Vectors)
library(ANCOMBC)
library(tibble)
library(MicrobiomeStat)
library(DAssemble)
library(cowplot)
library(ggplot2)
library(vegan)
library(plyr)
library(LOCOM)
library(DESeq2)
library(edgeR)
library(LOCOM)
library(ALDEx2)
library(ggplot2)
library(writexl)
library(stringr)

#BiocManager::install("ALDEx2")
library(ANCOMBC)
library(TreeSummarizedExperiment)

package_vec = c("reshape2", "ggplot2", "optparse", 
                "parallel", "stringi", "doParallel", "plyr", "tidyr", 'dplyr', 'vegan',
                'Maaslin2','ALDEx2')
loaded_packages <- intersect(package_vec, .packages())

library(here)
# Load Source Files
setwd(here::here())
#pkgmaker::source_files('./Codes/Real_Data/helper', '*.R')
```


## Data Loading and Preprocessing
The Preprocssing step is same with MaAsLin3 benchmark
https://github.com/WillNickols/maaslin3_benchmark/tree/main/HMP2/analysis/run_scripts

```{r}
# Manually define directories and parameters (for RStudio use)
workingDirectory <- "~/Desktop/DAssemble"
analysisDirectory <- "~/Desktop/DAssemble"

nCores <- 6  # Adjust based on your system
dataset <- "taxa"  # Choose 'taxa'
version <- 4  # Set MetaPhlAn version (3 or 4)

# Load MetaPhlAn data
if (version == 3) {
    taxa_table <- read.csv('~/Desktop/DAssemble/Codes/Dataset/metaphlan3_taxonomic_profiles.tsv', check.names = FALSE, sep = '\t')
    taxa_table <- taxa_table[grepl('\\|s__', taxa_table$`#SampleID`), ]  # Fixed escaping issue
    rownames(taxa_table) <- taxa_table$`#SampleID`
    taxa_table$`#SampleID` <- NULL
    taxa_table <- taxa_table * 100 # Convert to percentage
} else if (version == 4) {
    taxa_table <- read.csv('~/Desktop/DAssemble/Codes/Dataset/metaphlan4_taxonomic_profiles.tsv', skip = 1, check.names = FALSE, sep = '\t')
    colnames(taxa_table) <- gsub('_taxonomic$', '', colnames(taxa_table))
    taxa_table <- taxa_table[grepl('\\|t__', taxa_table$clade_name), ]  # Fixed escaping issue
    rownames(taxa_table) <- taxa_table$clade_name
    taxa_table$clade_name <- NULL
} else {
    stop("Invalid MetaPhlAn version")
}

prepare_metadata <- function(dataset_type) {
    metadata <- read.csv('~/Desktop/DAssemble/Codes/Dataset/hmp2_metadata_2018-08-20.csv', check.names = F)
    
    # Reorganize metadata table
    for (participant_id in unique(metadata$`Participant ID`)) {
        bmis <- metadata$BMI[metadata$`Participant ID` == participant_id]
        bmis <- bmis[!is.na(bmis)]
        metadata$BMI[metadata$`Participant ID` == participant_id] <- ifelse(length(bmis) > 0, mean(bmis), NA)
        
        smoke_status <- metadata$`smoking status`[metadata$`Participant ID` == participant_id]
        smoke_status <- smoke_status[!is.na(smoke_status)]
        metadata$`smoking status`[metadata$`Participant ID` == participant_id] <- ifelse(length(smoke_status) > 0, smoke_status[1], NA)
    }
    
    if (dataset_type == 'taxa') {
        metadata <- metadata[metadata$data_type == 'metagenomics',]
    } else {
        metadata <- metadata[metadata$data_type == 'metabolomics',]
    }
    rownames(metadata) <- metadata$`External ID`
    metadata <- metadata[,colSums(metadata == '', na.rm = T) != nrow(metadata)]
    keep_cols <- c('External ID', 'Participant ID', 'week_num', 'site_name', 'Age at diagnosis',
                   'Education Level', 'Occupation', 'consent_age', 'diagnosis',
                   colnames(metadata)[c(52:83, 85:111)], 'race', 'sex', 'BMI', 'reads_filtered')
    metadata <- metadata[,keep_cols]
    metadata <- metadata[,colSums(!is.na(metadata)) != 0]
    return(metadata)
}
metadata <- prepare_metadata('taxa')

# Calculate dysbiosis score
veg_dist_out <- vegdist(t(as.matrix(taxa_table)), method="bray")
veg_dist_out <- as.matrix(veg_dist_out)
dysbiosis_scores <- vector(length = nrow(veg_dist_out))
for (i in seq_along(rownames(veg_dist_out))) {
    sample_name <- rownames(veg_dist_out)[i]
    healthy_subset <- colnames(taxa_table)[colnames(taxa_table) %in% 
                                               rownames(metadata[metadata$week_num > 20 & 
                                                                     metadata$diagnosis == 'nonIBD' &
                                                                     metadata$`Participant ID` != metadata[sample_name,]$`Participant ID`,])]
    
    dysbiosis_scores[i] <- median(veg_dist_out[sample_name, healthy_subset])
}
names(dysbiosis_scores) <- rownames(veg_dist_out)

# Add the dysbiosis state to the metadata
dysbiosis_df <- data.frame(sample=names(dysbiosis_scores), dysbiosis_score=dysbiosis_scores)
metadata <- right_join(dysbiosis_df, metadata, by=c('sample'='External ID'))
metadata$dysbiosis_state <- metadata$dysbiosis_score > quantile(metadata$dysbiosis_score[metadata$diagnosis == 'nonIBD'], 0.9, na.rm=T)
metadata$dysbiosis_state <- ifelse(metadata$dysbiosis_state, paste0('dysbiosis_', metadata$diagnosis), 'none')
metadata$dysbiosis_state <- factor(metadata$dysbiosis_state, levels = c('none', 'dysbiosis_nonIBD', 'dysbiosis_UC', 'dysbiosis_CD'))

dysbiosis_df <- metadata[, c("sample", "dysbiosis_state")]
dysbiosis_df <- dysbiosis_df[order(dysbiosis_df$sample),]
dysbiosis_df <- dysbiosis_df[!duplicated(dysbiosis_df$sample),]

metadata <- prepare_metadata(dataset)
metadata <- right_join(dysbiosis_df, metadata, by=c('sample'='External ID'))

metadata$participant_id <- metadata$`Participant ID`
metadata$diagnosis <- factor(ifelse(metadata$diagnosis %in% c("CD", "UC"), "IBD", "nonIBD"), 
                             levels = c("nonIBD", "IBD"))


metadata <- metadata[!is.na(metadata$dysbiosis_state),]
rownames(metadata) <- metadata$sample

if (dataset == 'taxa') {
    if (version == 3) {
        taxa_table <- read.csv('~/Desktop/DAssemble/Codes/Dataset/metaphlan3_taxonomic_profiles.tsv',check.names = F, sep = '\t')
        taxa_table <- taxa_table[grepl('\\|s__', taxa_table$`#SampleID`) & 
                         !grepl('\\|t__', taxa_table$`#SampleID`) & 
                         taxa_table$`#SampleID` != 'UNCLASSIFIED',]
        rownames(taxa_table) <- taxa_table$`#SampleID`
        taxa_table$`#SampleID` <- NULL
        taxa_table['UNCLASSIFIED',] <- pmax(1 - colSums(taxa_table), 0)
        taxa_table <- taxa_table * 100 # Convert to percents to be consistent with v4
    } else if (version == 4) {
        taxa_table <- read.csv('~/Desktop/DAssemble/Codes/Dataset/metaphlan4_taxonomic_profiles.tsv', skip = 1, check.names = F, sep = '\t')
        
        # Reorganize taxa table
        colnames(taxa_table) <- gsub('_taxonomic$', '', colnames(taxa_table))
        # taxa_table <- taxa_table[taxa_table$clade_name == 'UNCLASSIFIED' | 
        #                              grepl('\\|t__', taxa_table$clade_name),]
        taxa_table <- taxa_table[grepl('\\|t__', taxa_table$clade_name) & 
                         taxa_table$clade_name != 'UNCLASSIFIED',]
        rownames(taxa_table) <- taxa_table$clade_name
        taxa_table$clade_name <- NULL
    } else {
        stop("--version not valid")
    }
    
} else {
    # Read in data
    mbx_table <- read.csv("~/Desktop/DAssemble/Codes/Dataset/intensities_hmp2.csv")
    annotations <- read.csv("~/Desktop/DAssemble/Codes/Dataset/annotations_hmp2.csv")
    
    if (all(abs(annotations$prev - rowMeans(!is.na(mbx_table))) < 0.001)) {
        mbx_table <- mbx_table[annotations$prim_feature == 'primary' & annotations$Metabolite != '',]
        rownames(mbx_table) <- paste0(annotations$Metabolite[annotations$prim_feature == 'primary' & annotations$Metabolite != ''], '_',
                                      annotations$HMDB.ID[annotations$prim_feature == 'primary' & annotations$Metabolite != ''], '_',
                                      annotations$Method[annotations$prim_feature == 'primary' & annotations$Metabolite != ''])
        mbx_table[is.na(mbx_table)] <- 0
    }
    taxa_table <- mbx_table
}

metadata <- metadata[metadata$sample %in% colnames(taxa_table),]
metadata <- metadata[as.numeric(mapvalues(metadata$sample, colnames(taxa_table), 1:ncol(taxa_table))),]

```

On the basis of same preprocessing as MaAsLin3, we further shorten feature names and subset the dataset to baseline to make model fitting easier.
```{r}
rownames(taxa_table) <- gsub("^.*\\|s__", "s__", rownames(taxa_table))

# Step 1: Extract baseline subjects from metadata
baseline_metadata <- metadata %>%
  distinct(participant_id, .keep_all = TRUE) %>%  
  filter(week_num < 3)  # Only subjects with week_num < 3


baseline_samples <- baseline_metadata$sample  
baseline_taxa_table <- taxa_table[, colnames(taxa_table) %in% baseline_samples, drop = FALSE]

cat("Dimensions of baseline_metadata:", dim(baseline_metadata), "\n")
cat("Subsetted Taxa Table Dimensions (samples × features):", dim(baseline_taxa_table), "\n")

taxa_table <- baseline_taxa_table
metadata <- baseline_metadata
```

## Run models

Here, we fit the two part model using MaAsLin2 and Logistic Regression. First, we fit the models separately and then combine the two-part model using the p-value combination method.

## Run MaAsLin2

```{r, echo=TRUE, results='hide'}
if (dataset == 'taxa') {
    if (version == 3) {
        tmp_fit_out <- paste0(gsub("/$", "", analysisDirectory), "/fit_out_MaAsLin2_3")
    } else {
        tmp_fit_out <- paste0(gsub("/$", "", analysisDirectory), "/fit_out_MaAsLin2_4")
    }
} else {
    tmp_fit_out <- paste0(gsub("/$", "", analysisDirectory), "/fit_out_MaAsLin2_mbx")
}

if (dataset == 'taxa') {
  fit_out <- Maaslin2::Maaslin2(taxa_table, metadata, 
                                min_abundance = 0, 
                                min_prevalence = 0, 
                                output = tmp_fit_out, 
                                min_variance = 0, 
                                normalization = 'TSS', 
                                transform = 'log', 
                                analysis_method = 'LM', 
                                fixed_effects = c('diagnosis', 'Antibiotics', 'consent_age'), 
                                save_scatter = FALSE, 
                                save_models = F, 
                                plot_heatmap = F, 
                                plot_scatter = F,
                                max_significance = 0.1, 
                                reference = 'diagnosis,nonIBD')$results
} else {
  fit_out <- Maaslin2::Maaslin2(taxa_table, 
                                metadata, 
                                min_abundance = 0, 
                                min_prevalence = 0, 
                                output = tmp_fit_out, 
                                min_variance = 0, 
                                normalization = 'NONE', 
                                transform = 'log', 
                                analysis_method = 'LM', 
                                fixed_effects = c('diagnosis', 'Antibiotics', 'consent_age'), 
                                save_scatter = FALSE, 
                                ave_models = F, 
                                plot_heatmap = F, 
                                plot_scatter = F,
                                max_significance = 0.1, 
                                reference = 'diagnosis,nonIBD')$results
}

unlink(tmp_fit_out, recursive = T)


fit_out <- data.frame(feature = fit_out$feature,
           metadata = fit_out$metadata,
           value = fit_out$value,
           coef = fit_out$coef,
           pval = fit_out$pval,
           qval = fit_out$qval,
           association = "abundance")

fit_out <- fit_out[fit_out$metadata == "diagnosis", ]
fit_out$qval <- p.adjust(fit_out$pval, method = "BH")


paras_CR <- fit_out
paras_CR <- paras_CR %>%
  dplyr::rename(
    coef_CR = coef, 
    pval_CR = pval,   
    qval_CR = qval
  ) %>%
  dplyr::mutate(
    metadata = "diagnosis" 
  ) %>%
  dplyr::select(
    feature,                
    coef_CR,              
    pval_CR,              
    qval_CR,
    metadata               
  )
#write.table(paras_CR, paste0(gsub("/$", "", analysisDirectory), "/results/", ifelse(dataset == 'taxa', '', 'mbx_'), "v", version, "_ibd_associations_MaAsLin2.tsv"), row.names = F, sep='\t')
```


## Run Logistic Regression
```{r, echo=TRUE, results='hide', warning=FALSE}
taxa_table <- as.data.frame(t(baseline_taxa_table))
metadata <- as.data.frame(baseline_metadata)
input_features <- taxa_table
input_metadata <- metadata
results_list <- list()

for (x in 1:ncol(input_features)) {
    
    feature_name <- colnames(input_features)[x] 
    featuresVector <- input_features[, x]  
    featuresVector <- as.factor(ifelse(featuresVector != 0, 1, 0))
    dat_sub <- data.frame(expr = featuresVector, metadata)
    formula <- as.formula("expr ~ diagnosis + consent_age + Antibiotics")
    
    fit <- tryCatch({
      glm(formula = formula, data = dat_sub, family = "binomial")
    }, error = function(err) {
      try(lme4::glmer(formula = formula, data = dat_sub, family = "binomial",
                      control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))))
    })
    
    if (!inherits(fit, "try-error")) {
        para <- as.data.frame(summary(fit)$coefficients[, c(1, 2, 4), drop = FALSE])
    } else {
        para <- as.data.frame(matrix(NA, nrow = 1, ncol = 3))
        colnames(para) <- c("Estimate", "Std. Error", "Pr(>|z|)")
        print(paste("Fitting problem for feature", feature_name, "returning NA"))
    }

    para$feature <- feature_name
    results_list[[feature_name]] <- para
}

results_df <- do.call(rbind, results_list)
results_df$variable <- rownames(results_df)  
diagnosis_subset <- results_df[grep("diagnosis", results_df$variable, ignore.case = TRUE), ]
diagnosis_subset$qval_LR <- p.adjust(diagnosis_subset$`Pr(>|z|)`, method = "BH")

paras_LR <- diagnosis_subset

paras_LR <- paras_LR %>%
  dplyr::mutate(
    metadata = "diagnosis"  
  ) %>%
  dplyr::select(
    feature,                
    coef_LR = Estimate,    
    pval_LR = `Pr(>|z|)`, 
    qval_LR,
    metadata              
  )

rownames(paras_LR) <- NULL

print(paras_LR)
#write.table(paras_LR, paste0(gsub("/$", "", analysisDirectory), "/results/", ifelse(dataset == 'taxa', '', 'mbx_'), "v", version, "_ibd_associations_LogisticRegression.tsv"), row.names = F, sep='\t')
```

## Combine results
After running MaAsLin2 and Logistic Regression, we use CC, MC and SC to combine p-values and calculate q-values based on combined p-values. We can use the DAssemble() from library(DAssemble) for this step.
```{r, warning=FALSE}
paras_CR <- paras_CR %>%
  dplyr::mutate(ID = gsub("\\.t__", "|t__", feature))  %>%  
  dplyr::rename(pvalue = pval_CR) 

paras_LR <- paras_LR %>%
  dplyr::mutate(ID = feature) %>% 
  dplyr::rename(pvalue = pval_LR) 

## MaAsLin2
dflist1 = list(paras_LR, paras_CR)
## MC
parasVanilla_MaAsLin2 = DAssemble(dflist1, combine.method = "minP", correction = "BH")
parasVanilla_MaAsLin2[1:10, ]
## CC
parasCCT_MaAsLin2 = DAssemble(dflist1, combine.method = "CCT", correction = "BH")
parasCCT_MaAsLin2[1:10, ]
## SC
parasStouffer_MaAsLin2 = DAssemble(dflist1, combine.method = "stouffer", correction = "BH")
parasStouffer_MaAsLin2[1:10,]

```


## Compare results

To compare Logistic Regression and models using different combination methods, we create plots to highlight key microbial taxa associated with IBD through combined abundance and prevalence modeling. Using MaAsLin2 + LR (SC) as a representative
method, we identified two major classes of IBD-associated taxa: 1) exhibiting both broad compositional shifts and presence-absence patterns (Fig. 6A); and 2) prevalence associations not detected by abundance modeling alone (Fig. 6B). 

## Example: MaAsLin2 + LR (SC)

```{r plot_pyramid, fig.width=12, fig.height=8, out.width="100%"}
threshold <- 0.25

# Extract significant species from non-ensemble model (MaAsLin2 - Original)
sig_non_ensemble <- paras_CR %>%
  filter(qval_CR < threshold) %>%
  dplyr::select(ID, qval_CR, pvalue, coef_CR) %>%
  mutate(Method = "Original")

# Extract significant species from ensemble models (CC, SC, MC)
sig_ensemble <- bind_rows(
  parasCCT_MaAsLin2 %>% mutate(Method = "CC"),
  parasStouffer_MaAsLin2 %>% mutate(Method = "SC"),
  parasVanilla_MaAsLin2 %>% mutate(Method = "MC")
) %>%
  filter(qval_BH < threshold)

sig_SC <- sig_ensemble %>%
  filter(Method == "SC") %>%
  arrange(qval_BH) 

common_species <- inner_join(sig_non_ensemble, sig_SC, by = "ID") %>%
  arrange(qval_BH) %>%
  slice_head(n = 10)  

# Select species significant in SC but NOT in non-ensemble
only_ensemble <- sig_SC %>%
  anti_join(sig_non_ensemble %>% filter(qval_CR < threshold), by = "ID") %>%
  arrange(desc(pvalue2))%>%  
  slice_head(n = 10)  

top_10 <- common_species %>%
  mutate(
    feature = ID,  
    pval_CR_log10 = -log10(pvalue2), 
    pval_LR_log10 = -log10(pvalue1),  
    coef_CR_sign = ifelse(coef_CR.x > 0, "+", "-"),  
    coef_LR_sign = ifelse(coef_LR > 0, "+", "-")  
  )

top_10 <- top_10 %>%
  mutate(feature = str_replace_all(ID, "s__|\\|t__.*", "")) %>%  
  mutate(feature = str_replace_all(feature, "_", " "))  

only_ensemble_10 <- only_ensemble %>%
  mutate(
    feature = ID,  
    pval_CR_log10 = -log10(pvalue2),  
    pval_LR_log10 = -log10(pvalue1),  
    coef_CR_sign = ifelse(coef_CR > 0, "+", "-"),  
    coef_LR_sign = ifelse(coef_LR > 0, "+", "-")  
  )

only_ensemble_10 <- only_ensemble_10 %>%
  mutate(feature = str_replace_all(ID, "s__|\\|t__.*", "")) %>%  
  mutate(feature = str_replace_all(feature, "_", " "))  

pval_range <- range(c(top_10$pval_CR_log10, only_ensemble_10$pval_CR_log10))
breaks_pretty_pval <- pretty(c(-max(pval_range), max(pval_range)), n = 5)

# Left Plot: Compare -log10(p-value) between MaAsLin2 and Logistic Regression (common species)
pyramid_plot_pval_1 <- ggplot(top_10) +
  geom_col(aes(x = feature, y = pval_CR_log10, fill = "MaAsLin2"), position = "identity") +
  geom_col(aes(x = feature, y = -pval_LR_log10, fill = "LR"), position = "identity") +
  geom_text(aes(x = feature, y = pval_CR_log10 + 0.3, label = coef_CR_sign), size = 7, hjust = 0.5) +  
  geom_text(aes(x = feature, y = -pval_LR_log10 - 0.3, label = coef_LR_sign), size = 7, hjust = 0.5) + 
  labs(y = "-log10(p-value)", x = "") +
  coord_flip() +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text.y = element_text(size = 11, hjust = 1, face = "italic", family = "serif"),
    axis.text.x = element_text(size = 10)
  ) +
  scale_fill_manual(values = c("MaAsLin2" = "darkblue", "LR" = "orange"), name = NULL) +
  scale_y_continuous(breaks = breaks_pretty_pval, labels = abs(breaks_pretty_pval), limits = range(breaks_pretty_pval))

# Right Plot: Compare -log10(p-value) between MaAsLin2 and Logistic Regression for species found in CC but not in MaAsLin2
pyramid_plot_pval_2 <- ggplot(only_ensemble_10) +
  geom_col(aes(x = feature, y = pval_CR_log10, fill = "MaAsLin2"), position = "identity") +
  geom_col(aes(x = feature, y = -pval_LR_log10, fill = "LR"), position = "identity") +
  geom_text(aes(x = feature, y = pval_CR_log10 + 0.3, label = coef_CR_sign), size = 7, hjust = 0.5) +  
  geom_text(aes(x = feature, y = -pval_LR_log10 - 0.3, label = coef_LR_sign), size = 7, hjust = 0.5) +  
  labs(y = "-log10(p-value)", x = "") +
  coord_flip() +
  theme_minimal() +
  theme(
    legend.position = "right",
    axis.title.x = element_text(size = 14),
    axis.title.y = element_blank(),
    axis.text.y = element_text(size = 11, hjust = 1, face = "italic", family = "serif"),
    axis.text.x = element_text(size = 10),
    legend.text = element_text(size = 10)
  ) +
  scale_fill_manual(values = c("MaAsLin2" = "darkblue", "LR" = "orange"), name = NULL) +
  scale_y_continuous(breaks = breaks_pretty_pval, labels = abs(breaks_pretty_pval), limits = range(breaks_pretty_pval))

sig_threshold <- -log10(0.05)

pyramid_plot_pval_1 <- pyramid_plot_pval_1 +
  geom_hline(yintercept = sig_threshold, linetype = "dashed", color = "red", size = 0.5) +  
  geom_hline(yintercept = -sig_threshold, linetype = "dashed", color = "red", size = 0.5)  

pyramid_plot_pval_2 <- pyramid_plot_pval_2 +
  geom_hline(yintercept = sig_threshold, linetype = "dashed", color = "red", size = 0.5) + 
  geom_hline(yintercept = -sig_threshold, linetype = "dashed", color = "red", size = 0.5)  


# Combine both plots
combined_plot <- plot_grid(
  pyramid_plot_pval_1 + 
    ggtitle("A. Abundance associations enhanced by prevalence modeling") + 
    theme(plot.title = element_text(size = 14, hjust = 0.9, margin = margin(b = 10)),
          plot.margin = margin(t = 10, r = 10, b = 15, l = 1)),  
  
  pyramid_plot_pval_2 + 
    ggtitle("B. Prevalence associations not detected by abundance modeling") + 
    theme(plot.title = element_text(size = 14, hjust = 0.7, margin = margin(b = 10)),
          plot.margin = margin(t = 10, r = 10, b = 1, l = 15)),  
  
  ncol = 2,
  align = "h",
  rel_widths = c(2, 2)
)

final_plot <- plot_grid(
  ggdraw() + draw_label(
    "",
    x = 0.5, hjust = 0.5, size = 50
  ),
  combined_plot,
  ncol = 1,
  rel_heights = c(0.01, 1.2)
)

print(final_plot)
```

In Panel A, taxa such as *Collinsella SGB14861* and *Clostridium sp AM49 4BH* show stronger associations when both abundance and prevalence are considered, suggesting their potential role in gut inflammation and metabolic alterations. Panel B reveals taxa like *Faecalibacterium intestinalis* and *Ruminococcus bicirculans*, whose prevalence associations were missed by abundance modeling alone, indicating their underrepresentation in IBD patients despite their potential protective roles. The detection of butyrate-producing bacteria, such as *Faecalibacterium intestinalis*, reinforces the known link between gut microbiota and intestinal health. 


## Other ensemble strategies

DAssemble can be applied to arbitrary pvalues. For example, p-values from an ANCOM-BC2 model can be combined with the LR p-values using the Cauchy Combination (CC) by using the following code:


## Run LinDA

```{r, echo=TRUE, results='hide'}
taxa_table <- baseline_taxa_table
metadata <- baseline_metadata

input_metadata <- metadata
input_features <- taxa_table

linda_result <- linda(
  feature.dat = input_features,  
  meta.dat = input_metadata,      
  formula = "~ diagnosis + Antibiotics + consent_age",
  is.winsor = FALSE,               
  zero.handling = "pseudo-count",  
  p.adj.method = "BH",            
  alpha = 0.05,                   
)

linda_result <- linda_result$output$diagnosisIBD
linda_result$qval <- p.adjust(linda_result$pval, method = "BH")

paras_LD <- linda_result

paras_LD <- paras_LD %>%
  tibble::rownames_to_column(var = "feature") %>% 
  dplyr::rename(
    coef_LD = log2FoldChange,  
    pval_LD = pvalue,           
    qval_LD = qval
  ) %>%
  dplyr::mutate(
    metadata = "diagnosis"      
  ) %>%
  dplyr::select(
    feature,   
    coef_LD,
    pval_LD,
    qval_LD,
    metadata
  )
print(paras_LD)
# write.table(paras_LD, paste0(gsub("/$", "", analysisDirectory), "/results/", ifelse(dataset == 'taxa', '', 'mbx_'), "v", version, "_ibd_associations_LINDA.tsv"), row.names = F, sep='\t')
```


```{r, warning=FALSE}
paras_LD <- paras_LD %>%
  dplyr::mutate(ID = gsub("\\.t__", "|t__", feature))  %>%  
  dplyr::rename(pvalue = pval_LD) 

dflist2 = list(paras_LR, paras_LD)
parasCCT_LinDA = DAssemble(dflist2, combine.method = "CCT", correction = "BH")
parasCCT_LinDA[1:10, ]
## Other ensemble methods:
## MinP
parasVanilla_LinDA = DAssemble(dflist2, combine.method = "minP", correction = "BH")
parasVanilla_LinDA[1:10, ]
## Stouffer
parasStouffer_LinDA = DAssemble(dflist2, combine.method = "stouffer", correction = "BH")
parasStouffer_LinDA[1:10,]

```

## Session information
```{r}
sessionInfo()
```



